Visualising CHD Pathway Inequalities by ethnicity - DRAFT

### 2024-03-12 findings.qmd
Further details added to the report

---
title: "Visualising CHD Pathway Inequalities by ethnicity - draft"
lang: en-GB
author: "Jacqueline Grout"
date: last-modified
date-format: "YYYY-MM-DD"
format:
  html:
    self-contained: true
    grid:
      sidebar-width: 200px
      body-width: 950px
      margin-width: 150px
      gutter-width: 1.5rem
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    fontcolor: black
    toc: true
    toc-location: left
    toc-title: Contents
    toc-depth: 3
editor: visual
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
editor_options: 
  chunk_output_type: console
---

@@

library(targets)
library(gt)
library(tidyverse)
library(grid)
library(gridExtra)
library(cowplot)
```

@@

In 2022, the Strategy Unit worked with the British Heart Foundation (BHF) to explore ways of visualising socio-economic inequalities as they emerge through the coronary heart disease (CHD) pathway. We made these visualisations available so that health and care staff can understand where on the pathway socio-economic inequalities emerge and at which points they are moderated or exacerbated. The report [^1]and web-based tool[^2] are available via the BHF and Strategy Unit websites.

@@

2.  Quantify and illustrate the inequalities by ethnicity over the disease progression and treatment pathway for CHD

@@

The term ‘inequities’ is used to describe unjustifiable differences in rates of access between
subgroups. An equity analysis must control for levels of need within each population subgroup.
Having done this, an equitable distribution of services is one where rates of access to a service or
population follow the distribution of need, such that a patient with a given level of need in one
subgroup has the same chance of accessing a service as their counterparts with a similar level of
need in other subgroups. This is the standard that the NHS seeks to achieve. Assessing equity is
challenging. Further detail about inequalities and inequities in healthcare can be found in a recent
Strategy Unit report[^3]

@@

metrics_table |>
    gt()|>
  cols_hide(metric_domain)|>
 # tab_stubhead(label = md("**Domain**"))|>
    tab_header(
    title = md("**Table 1** - *Coronary heart disease pathway metrics*")
  )|>
   tab_row_group(
     id="Need",
    label = md("**Need**"),
    rows = 1
  ) |>
  tab_row_group(
    id="Risk factors",
    label = md("**Risk factors**"),
    rows = 2
  ) |>
  tab_row_group(
    id="Risk factor identification",
    label = md("**Risk factor identification**"),
    rows = 3:7
  )|>
  tab_row_group(
    id="Primary prevention",
    label = md("**Primary prevention**"),
    rows = 8:12
  )|>
  tab_row_group(
    id="Disease identification",
    label = md("**Disease identification**"),
    rows = 13:15
  )|>
  tab_row_group(
    id="Secondary prevention",
    label = md("**Secondary prevention**"),
    rows = 16:23
  )|>
    tab_row_group(
    id="Tertiary prevention",
    label = md("**Tertiary prevention**"),
    rows = 24:29
  )|>
    tab_row_group(
    id="Intermediate outcome",
    label = md("**Intermediate outcome**"),
    rows = 30:32
  )|>
    tab_row_group(
    id="Full outcomes",
    label = md("**Full outcomes**"),
    rows = 33:36
  )|>
  row_group_order(groups=c("Need","Risk factors","Risk factor identification","Primary prevention","Disease identification","Secondary prevention","Tertiary prevention","Intermediate outcome","Full outcomes"))

#  cols_label(name_metric=md("**Metric**"))

    #domain_metric=md("**Domain**")
```

@@

Combining List size by lsoa with 2021 census by ethnicity

@@

#### Cluster 1 - Least diverse

@@

#### Cluster 2

@@

#### Cluster 3

@@

#### Cluster 4

@@

#### Cluster 5 - Most diverse

@@

@@

The charts below show the rates by cluster for each metric along the CHD pathway. The horizontal line is the overall rate. Explain how its calculated and how to interpret. Mention that confidence ontervals are also shown on the charts.

@@

```{r} tar_read(ci_iod_chart)}
```

@@

Confidence intervals are shown in grey. There is 95% confidence that the index of disparity is within this range.
:::

Calc of iod inc bootstrap of confidence intervals

@@

The charts below show the distribution of the patients in each region according to the cluster of their GP practice.

@@

Mention ICB charts are in appendix and refer to ICBs of interest under relevant region, hyperlink to appendix?? CIs of ICB????? should the CI be reduced?

@@

|-------------|---------------------|-----------|--------|-------------------|

@@

@@

@@

@@

@@

@@

CHD synthetic prevalence estimates data has been removed from Fingertips by OHID.

OHID have sent the archived data. This is for 7564 practices from 2015, of which 6297 are current practices.

Need to identify practices with missing data and work out how to get a prevalence for them.

Also have file from previous project - checked data is the same but need to check practices included - maybe there are more?

nearest neighbours to impute chd prevalence

<https://the-strategy-unit.github.io/data_science/blogs/posts/2024-01-17_nearest_neighbour.html>

@@

#### Which date should be used for the gp list size dataset?

### K-Medoids Cluster

@@

@@

Explain how these were done - bootstrap method etc

@@

### References

Confidence intervals vs statistical significance

<https://academic.oup.com/jrssig/article/16/4/20/7038025>

>>>>

---
title: "Visualising CHD Pathway Inequalities by ethnicity - DRAFT"
lang: en-GB
author: "Jacqueline Grout"
date: last-modified
date-format: "YYYY-MM-DD"
title-block-banner: "#f9bf07"
title-block-banner-color: "#333739"
format:
  html:
    self-contained: true
    grid:
      sidebar-width: 200px
      body-width: 950px
      margin-width: 150px
      gutter-width: 1.5rem
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    fontcolor: black
    toc: true
    toc-location: left
    toc-title: Contents
    toc-depth: 3
editor: visual
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
editor_options: 
  chunk_output_type: console

css: styles.css
---

@@

library(targets)
library(gt)
library(tidyverse)
library(grid)
library(gridExtra)

@@

In 2022, the Strategy Unit worked with the British Heart Foundation (BHF) to explore ways of visualising socio-economic inequalities as they emerge through the coronary heart disease (CHD) pathway. We made these visualisations available so that health and care staff can understand where on the pathway socio-economic inequalities emerge and at which points they are moderated or exacerbated. The report[^1] and web-based tool[^2] are available via the BHF and Strategy Unit websites.

@@

2.  Quantify and illustrate the inequalities by ethnicity over the disease progression and treatment pathway for CHD.

@@

The term ‘inequities’ is used to describe unjustifiable differences in rates of access between subgroups. An equity analysis must control for levels of need within each population subgroup. Having done this, an equitable distribution of services is one where rates of access to a service or population follow the distribution of need, such that a patient with a given level of need in one subgroup has the same chance of accessing a service as their counterparts with a similar level of need in other subgroups. This is the standard that the NHS seeks to achieve. Assessing equity is challenging. Further detail about inequalities and inequities in healthcare can be found in a previous Strategy Unit report[^3]

@@

metrics_table |>
    gt()|>
  cols_hide(metric_domain)|>
 # tab_stubhead(label = md("**Domain**"))|>
    tab_header(
    title = md("**Table 1** - *Coronary heart disease pathway metrics*")
  )|>
   tab_row_group(
     id="Need",
    label = md("**Need**"),
    rows = 1
  ) |>
  tab_row_group(
    id="Risk factors",
    label = md("**Risk factors**"),
    rows = 2
  ) |>
  tab_row_group(
    id="Risk factor identification",
    label = md("**Risk factor identification**"),
    rows = 3:7
  )|>
  tab_row_group(
    id="Primary prevention",
    label = md("**Primary prevention**"),
    rows = 8:12
  )|>
  tab_row_group(
    id="Disease identification",
    label = md("**Disease identification**"),
    rows = 13:15
  )|>
  tab_row_group(
    id="Secondary prevention",
    label = md("**Secondary prevention**"),
    rows = 16:23
  )|>
    tab_row_group(
    id="Tertiary prevention",
    label = md("**Tertiary prevention**"),
    rows = 24:29
  )|>
    tab_row_group(
    id="Intermediate outcome",
    label = md("**Intermediate outcome**"),
    rows = 30:32
  )|>
    tab_row_group(
    id="Full outcomes",
    label = md("**Full outcomes**"),
    rows = 33:36
  )|>
  row_group_order(groups=c("Need","Risk factors","Risk factor identification","Primary prevention","Disease identification","Secondary prevention","Tertiary prevention","Intermediate outcome","Full outcomes"))|>
  cols_label(name_metric="")

@@

In this analysis we used the k-medoids clustering method to group practices. Each practice was assigned to one of a small number of groups such that practices within the group have similar distributions of patients over ethnic groups. In order to obtain data on the ethnicity of a practice's registered population we imputed it based on the Lower Super Output Area (LSOA) of residence of GP registrants and the ethnicity distribution of patients living in those LSOAs using the Census 2021 data. Further detailed explanations of the method can be found in the appendix.

The clustering analysis resulted in 5 groups (clusters) of practices which are illustrated in the charts and map below.

@@

The median percentage of each ethnic group for each cluster was calculated and is presented in these charts to describe their respective diversity.

### Cluster 1 - Least diverse

@@

### Cluster 2

@@

### Cluster 3

@@

### Cluster 4

@@

### Cluster 5 - Most diverse

@@

The map below presents each GP practice coloured according to their assigned cluster. Cluster 1 (red) is the least diverse and cluster 5 (orange) is the most diverse.

@@

The charts below show the rates by cluster for each metric along the CHD pathway. The horizontal line is the overall rate. Explain how its calculated and how to interpret. Mention that confidence intervals are also shown on the charts.

@@

and explain what it means and what it is showing

```{r}
tar_read(ci_iod_chart)
```

@@

Confidence intervals are shown in grey. There is 95% confidence that the index of disparity is within this range. The confidence interval estimation methods are detailed in the appendix.
:::

@@

The charts below show the distribution of the patients in each NHS region according to the cluster of their GP practice.

@@

refer to ICBs of interest under relevant region

The appendix contains equivalent charts to those above, showing the Index of Disparity for each CHD pathway metric for each of the 42 ICBs.

@@

|---------------|---------------|---------------|---------------|---------------|

@@

Metrics taken from the QOF data via Fingertips have been extracted by GP practice and the counts
(numerator of the measure) from the calculations have been used.
Exception reporting metrics have used the Personalised Care Adjustment (PCA) data to obtain the
number of patients by GP practice for whom a PCA has been recorded for the relevant metric.
Possible reasons for a PCA include; Newly diagnosed/registered, Intervention is clinically
unsuitable, patient choice, did not respond to offers of care, service not available.

@@

SUS data was extracted using Transact-SQL and using relevant OPCS, ICD10, HRG, and Treatment
function codes as detailed in table 2 above.

@@

This data was supplied by NACR. The data wasn’t available by GP Practice, so an alternative method
was followed. Practices were allocated to a cluster and data was then supplied aggregated into clusters. In some cases, the data had to be redacted due to small numbers in the category. For the purposes of this analysis the redacted data item was reassigned an average value.
The true definition of eligible for rehabilitation is the HES admission and successful discharge of all
patients with CVD e.g., MI, revascularization, or HF. However, to link with the NACR only those
records entered into the database can be used so NACR use pseudo eligible as this is as close as
the audit can get, at a granular level, to eligible. It is though affected by areas without entry to the
database,
The data used relates to the 2023 calendar year, which is the most recent complete year, and
following changes in the way rehabilitation has been delivered due to the Covid pandemic, most
closely reflects the rehabilitation model now in place.

@@

The methodology used for this metric follows, as far as possible, that detailed by NHS Digital[^4]. The
readmissions counted relate to those who had an elective PCI / CABG in the period 1/3/2022 to
28/2/2023.

[^4]: <https://digital.nhs.uk/data-and-information/publications/statistical/ccg-outcomes-indicator-set/specifications/3.2-emergency-readmissions-within-30-days-of-discharge-from-hospital_1_4>

@@

ONS death records contain an encrypted HES ID which was used to link to outpatient, inpatient and
A&E HES records, from which the latest GP practice recorded (on a spine traced record) for the
patient was then assigned to the death record.

@@

Public Health England (PHE) CHD synthetic prevalence estimates data was previously available from OHID via Fingertips. This data has now been removed from Fingertips and archived, however OHID supplied it via email for this project following a request. This data is for 7564 practices from 2015, of which 6297 are current practices. In order to assign a prevalence for practices with missing data a nearest neighbours methodology was used to impute the prevalence by taking an average of the prevalence of each practice within a 1.5km radius of the practice without data. Further details, including code with sample data, for this methodology are available in a blog post on the Strategy Unit Data Science website[^5].

[^5]: <https://the-strategy-unit.github.io/data_science/blogs/posts/2024-01-17_nearest_neighbour.html>

### Ethnicity by GP practice

Ethnicity by GP practice was imputed based on the Lower Super Output Area (LSOA) of residence of GP registrants and the ethnicity distribution of patients living in these LSOAs using the Census 2021 data.

#### GP list size by LSOA

Since metric data is mainly from the 2022/23 year, the October 2022 list size data was taken from the NHS Digital Patients Registered at a GP practice dataset[^6] as this is the data closest to the mid point in the 22/23 year. The GP to ICB mapping file was also downloaded and joined to the list size to easily create analysis by ICB and region.

[^6]: <https://digital.nhs.uk/data-and-information/publications/statistical/patients-registered-at-a-gp-practice/october-2022>

@@

xxxxxxxxxxxxxx

explain how 2011 lsoa and 2021 lsoa mis match was handled

### K-Medoids Clustering

K-medoids is an established, and commonly used unsupervised machine learning technique. This clustering approach required data on the ethnicity of a practices registered population which was obtained as per the methodology explained above (Ethnicity by GP practice). The data used in the clustering was the % of patients in each ethnic group per practice, such that for each practice all the ethnic group %'s added to 100%.

The K-medoids clusterimg was performed using the {cluster} package[^7] in R using partitioning (clustering) of the data into k clusters "around medoids" (PAM), which is a more robust version of K-means.

[^7]: Maechler M, Rousseeuw P, Struyf A, Hubert M, Hornik K (2023). *cluster: Cluster Analysis Basics and Extensions*. R package version 2.1.6 — For new features, see the 'NEWS' and the 'Changelog' file in the package source), [https://CRAN.R-project.org/package=cluster](https://cran.r-project.org/package=cluster).

Clustering experiments were conducted using a variety of different combinations of ethnic breakdowns, such as 14 ethnic groups and 5 ethnic groups, as well as variations that excluded the % of white british patients or factored in age group. Techniques were also used to attempt to determine the optimal number of clusters such as the elbow method, gap plots, and scree plots. Consideration was given to combining some ethnic groups and in doing so a correlation matrix was produced to identify ethnic groups that commonly featured together. It was also important to seek a number of clusters that could be defined easily, where characteristics did not overlap each other so that resulting clusters were too similar to each other. It was also necessary to ensure that there was a sufficient number of practices in each cluster, too few practices in a cluster could result in issues with small amounts of metric data when calculating activity rates and consequently wide confidence intervals in the index of disparity calculations.

In conclusion, the final version selected and presented here used the percentage of the GP list size in each of the 14 census 2021 ethnic groups and five clusters was felt to give a sufficient number of describable clusters for which the index of disparity calculations could be performed. The chart below is an elbow plot generated for the 14 ethnic groups showing an elbow around 4 or 5 clusters.

```{r}
tar_read(elbow_plot)
```

Table 3 shows the number of GP practices assigned to each of the 5 clusters and the characteristics of the clusters are descirbed above in the main body of the report.

```{r}
final_data_full_cats_percent_5_clusters <- tar_read(final_data_full_cats_percent_5_clusters) |> as_tibble()

final_data_full_cats_percent_5_clusters |>
  group_by(cluster)|>
  count("practice_code")|>
  ungroup() |>
  select(cluster,n)|>
  gt()|>
    tab_header(title = md("**Table 3** - *Number of practices per cluster*"))|>
    cols_label(cluster=md("**Cluster**"),
             n=md("**Number of practices**"))
```

@@

remember...how , detail - allowing for NEED and how this is done

@@

Confidence intervals of 95% are presented in the national charts presented in this report. Confidence intervals have been used, rather than calculations of statistical significance as this presents a clearer picture of the likelihood of the disparity and due to concerns with the suitability in this case of statistical significance.

> "Practitioners need to be aware that statistical significance differs from practical importance in that statistical significance is highly dependent upon sample size. For a large sample, a statistically significant result is likely to be obtained even when the actual magnitude of an effect is small and of little or no practical importance. On the other hand, for a small sample, it is quite likely that insufficient evidence of a statistically significant result will be obtained – even when there is, indeed, an effect of practical importance." (Hahn, G. J., Doganaksoy, N., Meeker, W. Q., 2019)[^8]

[^8]: <https://academic.oup.com/jrssig/article/16/4/20/7038025>

In order to estimate the confidence intervals for the index of disparity, a bootstrap method was adopted, assuming independence, generating 1000 random samples for each metric for which an index of disparity was calculated. The 2.5th and 97.5th percentiles were then taken for each metric to give the upper and lower limits of the 95% confidence interval for each.

An explanation of bootstrap theory and confidence intervals is available in Practical Statistics for Data Scientists (Bruce, P. and Bruce, A., 2017)[^9]

[^9]: <https://www.oreilly.com/library/view/practical-statistics-for/9781491952955/>

@@

## R code on Github

All the code used in this project is available on the Strategy Unit GitHub[^10].

[^10]: <https://github.com/The-Strategy-Unit/bhf_chd-ethnicity>
